---
title: "Splines. Also, RSV"
output:
  ioslides_presentation: 
    css: presentation.css
  html_document:
    mathjax: default
  beamer_presentation: default
  slidy_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, fig.width=8, fig.height=4.5)
set.seed(0)

import::from(mgcv, gam)
import::from(plotly, plot_ly, ggplotly, add_lines, add_markers, config, add_ribbons)
import::from(ggplot2, ggplot, geom_line, geom_density, geom_point, geom_ribbon, geom_segment, coord_cartesian, theme_minimal, theme, element_text, scale_color_manual, scale_fill_manual, scale_linetype_manual, aes, labs, guides, guide_legend)
import::from(dplyr, group_by, ungroup, do, select, mutate, filter, rename)
import::from(plyr, ldply, llply, compact)
import::from(magrittr, `%>%`, `%<>%`)
import::from(RColorBrewer, brewer.pal)
import::from(htmlwidgets, saveWidget)
import::from(latex2exp, TeX)
```

# Agenda

 * Intro to splines
 * Modeling with splines
 * Outbreak outcome estimation
 * `pspline.inference` R package
 * Application to RSV

# Intro to splines

```{r, echo=FALSE}
# Plot piecewise continuous function over range. Range is a vector including points of discontinuity; for example range = c(1, 2, 3) means range is (1, 3) with a discontinuity at 2. 
add_func = function(range, dx, f, name=NA, breaks=TRUE) {
  data = seq(1:(length(range) - 1)) %>%
    ldply(function(idx) {
      x0 = range[idx]
      x1 = range[idx + 1]
      x = seq(x0, x1, by=dx)
      y = rep(f[[idx]](x), length.out=length(x)) # This rep here to catch the case when f(x) returns a single scalar
      x = c(x, x1)
      y = c(y, NA)
      data.frame(x=x, y=y)
    })

  plot = list(
    geom_line(data=data, aes(x=x, y=y, color=name), show.legend=TRUE)
  )
  if (breaks) {
    plot %<>% c(list(
      geom_point(data=data %>% filter(x %in% range), aes(x=x, y=y, color=name))
    ))
  }
  plot
}

base_plot = function(data=NULL) {
  palette1 = rev(brewer.pal(5, "YlGnBu"))[2:4]
  palette2 = rev(brewer.pal(9, "YlGnBu"))[2:6]

  aesthetics = list(
    "true" = list(
      label = "True",
      color = "#FF0000",
      linetype = "12"
    ),
    "obs" = list(
      label = "Observed",
      color = "#C0C0C0"
    ),
    "est" = list(
      label = "Estimated",
      color = "#000000",
      linetype = "solid"
    ),
    "est1" = list(
      label = "Estimate #1",
      color = palette2[1],
      linetype = "solid"
    ),
    "est2" = list(
      label = "Estimate #2",
      color = palette2[2],
      linetype = "solid"
    ),
    "est3" = list(
      label = "Estimate #3",
      color = palette2[3],
      linetype = "solid"
    ),
    "est4" = list(
      label = "Estimate #4",
      color = palette2[4],
      linetype = "solid"
    ),
    "est5" = list(
      label = "Estimate #5",
      color = palette2[5],
      linetype = "solid"
    ),
    "est-95-cl" = list(
      label = "Estimated (95% CL)",
      fill = palette1[1]
    ),
    "est-dist" = list(
      label = "Estimated distribution",
      fill = palette1[2]
    ),
    "y" = list(
      label = TeX("f(x)"),
      color = palette1[1],
      linetype = "solid"
    ),
    "y1" = list(
      label = TeX("f'(x)"),
      color = palette1[2],
      linetype = "solid"
    ),
    "y2" = list(
      label = TeX("f''(x)"),
      color = palette1[3],
      linetype = "solid"
    ),
    "yhat" = list(
      label = TeX("\\hat{f}(x)"),
      color = palette1[1],
      linetype = "solid"
    ),
    "yhat1" = list(
      label = TeX("\\hat{f}'(x)"),
      color = palette1[2],
      linetype = "solid"
    ),
    "yhat2" = list(
      label = TeX("\\hat{f}''(x)"),
      color = palette1[3],
      linetype = "solid"
    )
  )
  
  makeScale = function(scale, aesName) {
    aesValid = llply(aesthetics, function(aes) {
      if (!is.null(aes[[aesName]])) {
        aes
      }
    }) %>% compact()
    aesValues = llply(aesValid, function(aes) aes[[aesName]])
    aesLabels = llply(aesValid, function(aes) aes$label)
    
    scale(
      name=NULL,
      breaks=names(aesValues),
      values=unlist(aesValues),
      labels=aesLabels
    )
  }
  
  ggplot(data=data) + 
    makeScale(scale_color_manual, "color") +
    makeScale(scale_fill_manual, "fill") +
    makeScale(scale_linetype_manual, "linetype") +
    theme_minimal() + 
    theme(
      legend.text=element_text(size=12),
      legend.position="bottom"
    ) +
    labs(x=NULL, y=NULL)
}

show_plot = function(plot) {
  plot
}
```

## Splines: what are they?

A *spline* is a *piecewise polynomial* on $[a, b] \subset \mathbb{R}$. The points at which the polynomial *segments are joined* are called *knots*.

```{r fig.height=2.25}
(base_plot() +
  add_func(c(0, 1, 2, 3), 0.001, c(
    function(x) x, 
    function(x) (17/16 - (x-1.25)**2),
    function(x) 0.25
  ), name="y")
) %>% show_plot()
```


If $P_i$ are *polynomials of degree ≤ m*, then the *spline* S has *degree m* and order m+1.

A spline is *continuous in all derivatives* everywhere *except* at its *knots*. At its knots a spline can have *continuity order* ranging from -1 (*no continuity*) through m-1.

## Spline degree 1, continuity -1

```{r}
f = c(
  function(x) 1.25 - x, 
  function(x) 1 - x / 2
)

(base_plot() + 
  add_func(c(0, 1, 2), 0.001, f, name="y")
  # add_func(c(0, 1, 2), 0.001, f2, name="dy/dx") %>%
) %>% show_plot()
```

## Spline degree 1, continuity 0

```{r}
f = c(
  function(x) 1.5 - x, 
  function(x) 1 - x/2
)
df = c(
  function(x) -1, 
  function(x) -1/2
)

(base_plot() +
  add_func(c(0, 1, 2), 0.001, f, name="y") +
  add_func(c(0, 1, 2), 0.001, df, name="y1", breaks=FALSE)
) %>% show_plot()
```


## Spline degree = 2, continuity -1

```{r}
f = c(
  function(x) x**2, 
  function(x) 2 - (x - 2)**2 / 2
)

(base_plot() +
  add_func(c(0, 1, 2), 0.001, f, name="y")
) %>% show_plot()
```

## Spline degree 2, continuity = 0

```{r}
f = c(
  function(x) 0.5 + x**2, 
  function(x) 2 - (x - 2)**2 / 2
)
df = c(
  function(x) x*2, 
  function(x) 2-x
)

(base_plot() +
  add_func(c(0, 1, 2), 0.001, f, name="y")+
  add_func(c(0, 1, 2), 0.001, df, name="y1", breaks=FALSE)
) %>% show_plot()

```

## Spline degree 2, continuity 1

```{r}
f = c(
  function(x) x**2,
  function(x) 2 - (x - 2)**2
)
df = c(
  function(x) 2 * x,
  function(x) -2 * (x - 2)
)
d2f = c(
  function(x) 2, 
  function(x) -2
)

(base_plot() +
  add_func(c(0, 1, 2), 0.001, f, name="y") +
  add_func(c(0, 1, 2), 0.001, df, name="y1", breaks=FALSE) +
  add_func(c(0, 1, 2), 0.001, d2f, name="y2", breaks=FALSE)
) %>% show_plot()
```

# Spline modeling

## Spline modeling

Goal: fit a spline to observations. *Cubic splines* are commonly used.

In a simple spline model, *quality of fit* is measured by *mean square error* (MSE).

```{r}
deriv_name = Vectorize(function(d) {
  if (d == 0) {
    "yhat"
  } else {
    sprintf("yhat%d", d)
  }
})

add_psplines = function (x, y, k, m.fit, m.penalty, cyclic=FALSE, show.knots=TRUE, deriv=0) {
  data.obs = data.frame(x=x, y=y)
  if (cyclic) {
    model = gam(y ~ s(x, k=k, bs="cp", m=c(m.fit, m.penalty)), family=gaussian, data=data.obs)
  } else {
    model = gam(y ~ s(x, k=k, bs="ps", m=c(m.fit, m.penalty)), family=gaussian, data=data.obs)
  }
  
  data.pred = data.obs %>% 
    select(x) %>%
    mutate(y = predict(model, data.frame(x=x)), d=0)
  
  if (deriv > 0) {
    for (idx in 1:deriv) {
      data.pred %<>%
        filter(d == idx - 1) %>%
        do((function(df) {
          dy = diff(df$y) / diff(df$x)
          if (cyclic) {
            dy = (c(tail(dy, 1), dy) + c(dy, head(dy, 1))) / 2
          } else {
            dy = (c(head(dy, 1), dy) + c(dy, tail(dy, 1))) / 2
          }
          data.frame(x=df$x, y=dy, d=idx)
        })(.)) %>% 
        rbind(data.pred)
    }
  }

  knots = model$smooth[[1]]$knots
  if (!cyclic) {
    knots = knots[(m.fit+2):(length(knots)-m.fit-1)]
  }

  knots %<>% data.frame(x=.)
  knots = model %>% 
    predict(knots) %>% 
    data.frame(y=.) %>%
    cbind(knots)
  
  c(
    geom_point(data=data.obs, aes(x=x, y=y, color="obs")),
    geom_line(data=data.pred, aes(x=x, y=y, group=d, color=deriv_name(d))),
    geom_point(data=knots, aes(x=x, y=y, color=deriv_name(0)))
  )
}
```

## Linear spline model, 3 knots

```{r}
set.seed(0)
f = function(x) (1 - (2 * x - 1/2)**2)/2
x = seq(0, 1, by=0.01)
y = rnorm(length(x), mean=f(x), sd=0.025)

(base_plot() +
  add_psplines(x, y, 3, 0, 0)
) %>% show_plot()
```

## Linear spline model, 10 knots

```{r}
(base_plot() +
  add_psplines(x, y, 10, 0, 0)
) %>% show_plot()

```

## Linear spline model, 20 knots

```{r}
(base_plot() +
  add_psplines(x, y, 20, 0, 0)
) %>% show_plot()
```

## Overfitting and penalties

Spline fitting with *MSE is sensitive to number of knots*. Too *many knots* lead to *overfitting*.

One solution to this problem is to include *smoothness* in *measure of fit*; this is known as *"penalty"*.

Most common penalty is *integral of second derivative squared*, also known as 2nd order penalty.

*Splines* used for fitting *with a penalty* term are known as *penalized splines*, or *P-splines*.

## Linear P-spline model, 20 knots

```{r}
(base_plot() +
  add_psplines(x, y, 20, 0, 2)
) %>% show_plot()
```

## Spline modeling: cyclic time

Splines *aren't smooth* at the “ends” of a *time cycle*.

```{r}
set.seed(0)
f = function(x) cos(x)
x = seq(0, 2 * pi, by=0.01)
y = rnorm(length(x), mean=f(x), sd=0.025)

(base_plot() +
  add_psplines(x, y, 7, 2, 2, deriv=1)
) %>% show_plot()
```

## Spline modeling: cyclic time

That is what *cyclic splines* are used for.

```{r}
(base_plot() +
  add_psplines(x, y, 4, 2, 2, cyclic=TRUE, deriv=1)
) %>% show_plot()

```

# Outbreak outcome estimation with P-splines

## 1. Define outcome measure

Onset = time corresponding to 2.5% of total cases

```{r}
set.seed(0)
start = 20
end = 40
peak = 300

t.obs = seq(1, 52)
dt = 0.01

t = seq(min(t.obs) - 0.5, max(t.obs) + 0.52 - dt, by=dt)
cases0 = round((1 - cos((t - start) / (end - start) * 2 * pi)) / 2 * peak) * (t >= start) * (t <= end)
data0 = data.frame(time=t, cases=cases0)
m0 = gam(cases ~ s(x=time, k=20, bs="ps", m=3), family=poisson, data=data0)
cases.true = predict(m0, data.frame(time=t), type="response")
data.true = data.frame(time=t, cases=cases.true)

threshold = 0.025

cum.frac.true = cumsum(cases.true) / sum(cases.true)
t.onset.true = t[cum.frac.true <= threshold]
cases.onset.true = cases.true[cum.frac.true <= threshold]
data.onset.true = tail(data.frame(time=t.onset.true, cases=cases.onset.true), 1)

(base_plot() +
  geom_line(data=data.true, aes(x=time, y=cases, color="true"), size=1) + 
  geom_segment(data=data.onset.true, aes(x=time, xend=time, y=0, yend=cases, color="true"), size=1) + 
  labs(x="Week", y="Cases")
) %>% show_plot()
```

## 2. Fit a P-spline model

```{r}
cases.true0 = predict(m0, data.frame(time=t.obs), type="response")
cases.obs = rpois(length(cases.true0), cases.true0)
data.obs = data.frame(time=t.obs, cases=cases.obs)

m = gam(cases ~ s(x=time, k=8, bs="cp", m=3), family=poisson, data=data.obs)
cases.pred = predict(m, data.frame(time=t), type="response")
data.pred = data.frame(time=t, cases=cases.pred)
  
(base_plot() +
  geom_line(data=data.true, aes(x=time, y=cases, color="true"), size=1) + 
  geom_point(data=data.obs, aes(x=time, y=cases, color="obs")) +
  geom_line(data=data.pred, aes(x=time, y=cases, color="est"), size=0.5) + 
  labs(x="Week", y="Cases")
) %>% show_plot()
```

## 3. Sample model parameters

```{r estSmall, cache=TRUE}

library(pspline.inference)

sample_name = function(n) {
  sprintf("est%d", n)
}

n=5
cases.samples = pspline.sample.timeseries(m, data.frame(time=t), pspline.outbreak.cases, samples=n)
```

```{r} 
(base_plot(cases.samples) +
  geom_line(data=data.true, aes(x=time, y=cases, color="true"), size=1) + 
  geom_point(data=data.obs, aes(x=time, y=cases, color="obs")) +
  geom_line(aes(x=time, y=cases, group=pspline.sample, color=sample_name(pspline.sample)), size=0.5) + 
  labs(x="Week", y="Cases")
) %>% show_plot()
```

## 3. Sample model parameters

```{r} 
zoom1 = coord_cartesian(x=c(20, 23.5), y=c(0, 80))

(base_plot(cases.samples) +
  geom_line(data=data.true, aes(x=time, y=cases, color="true"), size=1) + 
  geom_point(data=data.obs, aes(x=time, y=cases, color="obs")) +
  geom_line(aes(x=time, y=cases, group=pspline.sample, color=sample_name(pspline.sample)), size=0.5) + 
  labs(x="Week", y="Cases") +
  zoom1
) %>% show_plot()
```

## 4. Sample outcome measure

```{r}
onset.samples = cases.samples %>% 
  group_by(pspline.sample) %>%
  do((function(data){
    cases.frac = cumsum(data$cases) / sum(data$cases)
    data.frame(
      pspline.sample = tail(data$pspline.sample, 1),
      onset = tail(data$time[cases.frac <= threshold], 1),
      cases = tail(data$cases[cases.frac <= threshold], 1)
    )
  })(.)) %>%
  ungroup()

#(base_plot(cases.samples) +
#  geom_line(data=data.true, aes(x=time, y=cases, color="true"), size=1) + 
#  geom_point(data=data.obs, aes(x=time, y=cases, color="obs")) +
#  geom_line(aes(x=time, y=cases, group=pspline.sample, color=sample_name(pspline.sample)), size=0.5) +
#  geom_segment(data=onset.samples, aes(x=onset, xend=onset, y=0, yend=cases, group=pspline.sample, color=sample_name(pspline.sample)), size=0.5) + 
#   labs(x="Week", y="Cases")
#) %>% show_plot()
```

```{r}
(base_plot(cases.samples) +
  geom_line(data=data.true, aes(x=time, y=cases, color="true"), size=1) + 
  geom_point(data=data.obs, aes(x=time, y=cases, color="obs")) +
  geom_line(aes(x=time, y=cases, group=pspline.sample, color=sample_name(pspline.sample)), size=0.5) +
  geom_segment(data=onset.samples, aes(x=onset, xend=onset, y=0, yend=cases, group=pspline.sample, color=sample_name(pspline.sample)), size=0.5) + 
  labs(x="Week", y="Cases") +
  zoom1
  
) %>% show_plot()
```

## 5. Infer outcome distribution

```{r estLarge, cache=TRUE}
n.2k = 2000
onset.samples.2k = pspline.sample.scalars(m, data.frame(time=t), pspline.outbreak.thresholds(onset=0.025), samples=n.2k)
cases.est.2k = pspline.estimate.timeseries(m, data.frame(time=t), pspline.outbreak.cases, samples=n.2k)
```

```{r density}
(base_plot(cases.samples) +
  geom_ribbon(data=cases.est.2k, aes(x=time, ymin=cases.lower, ymax=cases.upper, fill="est-95-cl"), color=NA) +
  geom_density(data=onset.samples.2k, aes(x=onset, y=..density..*5, fill="est-dist"), color=NA, trim=TRUE) +
  geom_line(data=data.true, aes(x=time, y=cases, color="true"), size=1) + 
  geom_segment(data=data.onset.true, aes(x=time, xend=time, y=0, yend=cases, color="true"), size=1) + 
  geom_point(data=data.obs, aes(x=time, y=cases, color="obs")) +
  labs(x="Week", y="Cases") +
  guides(fill=guide_legend(), color="none", linetype="none") +
  zoom1
) %>% show_plot()
```

# `pspline.inference`

## `pspline.estimate.scalars`

* Set up a model:

```{r echo=TRUE, eval=FALSE}
model <- gam(..., data)
```

* Provide outcome calculation from data:

```{r echo=TRUE, eval=FALSE}
calc.outcome <- function(data) { ... }
```

* Get your estimates:

```{r echo=TRUE, eval=FALSE}
outcome.estimate <- pspline.estimate.scalars(
  model, data, calc.outcome
)
```

## `pspline.validate.scalars`

* Generate simulated truth
```{r echo=TRUE, eval=FALSE}
gen.truth <- function() { ... }
```

* Sample observations from truth
```{r echo=TRUE, eval=FALSE}
gen.observations <- function(truth) { ... }
```

* Set up model for observations
```{r echo=TRUE, eval=FALSE}
gen.model <- function(observations) { ... }
```

## `pspline.validate.scalars`

* Validate your outcome estimation
```{r echo=TRUE, eval=FALSE}
validation <- pspline.validate.scalars(
  gen.truth, n.truth, 
  gen.observations, n.observations, 
  gen.model, calc.outcome
)
```

# RSV

## Motivation

*RSV* is one of the top infectious causes of *infant hospitalizations and mortality* in developed countries.

Effective *prophylaxis (palivizumab) is expensive*, and therefore only recommended during RSV season.

*Prophylaxis* guidelines (by American Academy of Pediatrics) use the *same schedule for all states* except for Florida.

Is that schedule a good match for the actual RSV season timing in CT?

## Motivation

```{r}
htmltools::tags$img(
  src="coverageByRegionAAPInsight-Standalone.pdf",
  style="width: 768px;"
)
```

## Questions

How many RSV cases occur during the AAP-recommended prophylaxis protection window?

How much can we increase that by adjusting the window based on state-level and county-level data?

## Approach

If prophylaxis lasts from $T_\text{start}$ to $T_\text{end}$, protected fraction is defined as:

\[
\begin{align}
\text{protected } & \text{fraction}(T_\text{start}, T_\text{end}) = \\
&\frac{\text{RSV cases occurring beteween } T_\text{start} \text{ and } T_\text{end}}{\text{all RSV cases}}
\end{align}
\]

Using `pspline.inference`:

1. Estimate protected fraction for AAP prophlaxis guidelines.
2. Estimate protected fraction for alternate prophylaxis schedule --- same duration, but lined up with RSV season and aligned to useful calendar intervals

## Results

Protected fraction of *AAP*-recommended prophylaxis is *94.08% statewide* (95% CI: 93.70 -- 94.42%)

Aligning to RSV season *by county or statewide* and rounding prophylaxis window to *1 or 2 weeks* increases protected fraction by *~0.75-2%*. 

Least increase: 1-week rounding, season alignment by county = 94.81% statewide protected fraction (95% CI: 94.47 -- 95.12%).

Rounding schedule to 4 weeks gives no benefit over AAP. (AAP is as good as it gets for month-based schedule in CT.)

Adjusting for year-to-year variation in RSV season timing doesn't add any benefit.

## Conclusions

*Potential ~1% gain in protection* by adjusting prophylaxis schedule to match local (state or county) season timing (rounded to 1- or 2-week intervals).

*Need cost-benefit analysis* to weigh 1% improvement against implementation complexity.

Analysis method *generalizes to other states*.

`pspline.inference` *generalizes to other outcomes*.

